Multi-output Bus Travel Time Prediction\\ with Convolutional LSTM Neural Networks

### 2017-11-21 paper/paper.tex
Fix math notation.
Final draft

\address{
  \vspace{.5em}
  \begin{minipage}{.5\textwidth}
    \center
    Department of Management Engineering \\
    Technical University of Denmark, 2800 Kongens Lyngby 
  \end{minipage}% This must go next to `\end{minipage}`
  \begin{minipage}{.5\textwidth}
    \center
    Public Transport Movia \\
    Gammel Køge Landevej 3, 2500 Valby 
  \end{minipage}
  \vspace{-2.5em}
}

\begin{abstract}
Accurate and reliable predictions for travel times in public transport networks is essential for delivering an attractive service.
This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers. The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services. We find that the proposed model significantly outperforms all the other methods, and is able to detect small irregular peeks in bus travel time very quickly.
\end{abstract}

@@

In this paper we present a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers. The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas. The method is empirically evaluated and compared to other popular approaches for link travel time prediction, including the currently model deployed in production by the public transport authority for the Greater Copenhagen Area: \emph{Movia}. Furthermore the method is compared to Google Traffic, a popular online service for travel time prediction.

@@

Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies for capturing \emph{changes} in the travel time, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Yo2010,Bai2015}. The Kalman-filter models still have limited options for capturing and forecasting the complex dynamics of travel and dwell time in a metropolitan bus system. E.g., the Kalman-filter's state is only directly accessible for the leading time-step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time-steps. This is substantiated by \cite{Lin2013} and \cite{Kumar2014}, who find artificial neural networks (ANN) to outperform Kalman-filter models. However, the computational challenges of fully connected ANNs is also limiting the depth of the network and thus how complex patterns it can learn to recognize. This has sparked the interest for studying composite or hybrid models. \citet{Bai2015} uses a two-stage approach by combining offline ANN-models, with an adaptable/online Kalman-filter to yield a dynamic model. The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey, and is still not able to recognize long distance patterns. The proposed model in this work uses \emph{Long short-term memory} cells (LSTM), and differs from existing research in bus link travel time prediction by combining the capability for persisting state-space over multiple time-steps, while allowing the deep neural network to be feasibly trainable.

@@

\section{Convolutional LSTM neural networks}
\label{sec:convlstm}
A Long short-term memory (LSTM) neural network is a special type of Recurrent Neural Network~(RNN) which has proven robust for capturing long-distance dependencies \cite{Lstm1,Lstm2}. The important feature of a LSTM network is its capability to persist cell state, $\matr{c}_t$, from previous observations across sequences of input (e.g.\ time), but also eliminate information which is considered irrelevant. To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}. Each gate yields a state value at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.

\begin{equation}
\begin{aligned}
  \matr{i}_t &= \sigma \left( \matr{W}^i \matr{x}_t + \matr{R}^i \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
  \matr{f}_t &= \sigma \left( \matr{W}^f \matr{x}_t + \matr{R}^f \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c \matr{x}_t + \matr{R}^c \matr{h}_{t-1} + \matr{b}_c \right) \\
  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} \matr{x}_t + \matr{R}^\mathit{o} \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
\end{aligned}
\label{eq:lstm}
\end{equation}
\vspace{.5em}

@@

Convolutional Neural Networks~(CNN) on the other hand have been widely used for capturing spacial relationships, e.g.\ importance of neighboring pixels in an image. As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuse the same weights for producing several outputs. Instead of considering the entire input-vector, only a fixed-size window, or \emph{convolution}, around each input is considered. The weights are therefore referred to as the \emph{filters} or \emph{kernels} of the layer. \Cref{fig:conv} illustrates a single convolutional filter of size $3$ being applied to 1-dimensional data.

@@

\citet{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}. Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions of the LSTM cf.~\cref{eq:convlstm}, where $*$ denotes the convolution operator.
\begin{equation}
\begin{aligned}
  \matr{i}_t &= \sigma \left( \matr{W}^i * \matr{x}_t + \matr{R}^i * \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
  \matr{f}_t &= \sigma \left( \matr{W}^f * \matr{x}_t + \matr{R}^f * \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c * \matr{x}_t + \matr{R}^c * \matr{h}_{t-1} + \matr{b}_c \right) \\
  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} * \matr{x}_t + \matr{R}^\mathit{o} * \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
\end{aligned}
\label{eq:convlstm}
\end{equation}
\vspace{.5em}

@@

Thus convolutional filters are applied to each input, at each time-step, to the respective LSTM-cell, and also between LSTM-cells in the state-transition. Since the time-steps are one-dimensional~(i.e.\ link travel times across links), the filters are also one-dimensional. In each of the two blocks, the \emph{ConvLSTMs} are arranged with filter sizes of respectively $10\times1$ and $5\times1$ for each of the layers in the block. This size is used both for the \emph{input-to-state} and \emph{state-to-state} convolutional filters. Finally each \emph{ConvLSTM} layer 64~outputs, yielding a total of 512 convolutional filters.

@@

For the \emph{ConvLSTM} model to be able to capture the desired spatial-–temporal patterns, the input data must be arranged in a suitable manner: I.e.\ in $N$ samples, each with a window of the $w$ lagging time-steps $t-w+1, \ldots, t$, and each time-step with $u$ link travel times $1, \ldots, u$ cf.~\Cref{fig:data_shape}.

@@

\subsection{Detrending}
Urban bus travel times vary throughout the time of the day, and the day of the week due to \emph{recurring congestion}. In order to reduce the need for the network to learn this recurring variation, link travel for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern. Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.~\cref{eq:normalization}.

@@

Given the time resolution, we set the fixed windows size, $w = 32$, equivalent of 8 hours, to allow pattern in the morning peek affect patterns in the afternoon peek. We set $k = 3$ to allow predictions of up to 45 minutes into the future.

@@

\section{Results and discussion}
\label{sec:results}
The performance of our proposed model, based on \emph{ConvLSTM}, for link travel time prediction is compared against several other baseline models and services using the presented evaluation:
\begin{enumerate}
   \item a naïve historical average model, i.e.\ equivalent of just predicting the normalized value, $\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}$ 
   \item the current traffic prediction model currently deployed by Movia.
   \item a pure LSTM-model for link travel time prediction, i.e.\ without applying convolutional filters in state transitions.
   \item travel time predictions from Google Traffic (part of Google Maps).
\end{enumerate}

@@

\subsection{Current model}
Measurements for the currently deployed bus prediction model were collected at a 5-minute frequency using a non-publicly accessible endpoint at the transport authority. The model is based on an historical average model, but has rule based mechanism on top, which can override or adjust the historical link travel times. For instance it will assume that a delayed vehicle can recover (partially or completely) from its delay by traversing links faster. Of course such an assumption can be problematic in an urban area with lots of external traffic effects.

@@

\subsection{Google Traffic}
Measurements from the Google Traffic model were collected using the Google Maps Distance Matrix API~\citep{GMaps_DistanceMatrixAPI}. Google uses crowd-sourced road congestion data collected from smart-phones with the \emph{Google Maps App} installed \citep{GMaps_Crowdsourcing}. While the exact model powering the service is not publicly described in detail, the documentation states that ``that the returned \emph{duration in traffic} should be the best estimate of travel time given what is known about both historical traffic conditions and live traffic''. And further, ``that live traffic becomes more important the closer the departure time is to now'' \citep{GMaps_DistanceMatrixAPI}.

@@

Another important aspect is that the Google Traffic model is primary designed for estimating car travel times, and can thus be biased away from the bus travel times used in this experiment. Since we only consider link travel time, and collection data for each link individually, the bus dwell time will not be an issue, as it is not included in either measurement.

@@

Although the advantage might seem small it is emphasized that evaluation measurements are averaging their response, and thus the increased accuracy can be much higher on individual journeys, especially if they experience very irregular travel times. To investigate this, we focus our analysis on periods when the transport system is most vulnerable, and even small changes in regularity can propagate, since recovery is not an option, i.e.\ morning and afternoon peeks.

\begin{table}[!ht]
  \center
  \begin{tabular}{l|rrr}
  Model &  RMSE &  MAE &  MAPE \\
  \hline
  \hline
  Historical Average &        6.40 &       5.57 &     10.62 \% \\ \hline
  Current Model      &        6.69 &       5.88 &     11.22 \% \\ \hline
  Pure LSTM          &        0.00 &       0.00 &      0.00 \% \\ \hline
  Google Traffic     &        5.25 &       4.62 &      9.17 \% \\ \hline
  ConvLSTM           &        2.64 &       2.09 &      4.04 \% \\ \hline
  \end{tabular}
  \caption{Results: Morning peek (7--9)}
  \label{tab:morning_peek}

@@

\begin{table}[!ht]
  \center
  \begin{tabular}{l|rrr}
  Model &  RMSE &  MAE &  MAPE \\
  \hline
  \hline
  Historical Average &    5.90 &   4.65 &  8.28 \% \\ \hline
  Current Model      &    6.28 &   5.20 &  9.37 \% \\ \hline
  Pure LSTM          &    0.00 &   0.00 &  0.00 \% \\ \hline
  Google Traffic     &    4.16 &   3.34 &  6.21 \% \\ \hline
  ConvLSTM           &    3.79 &   3.02 &  5.61 \% \\ \hline
  \end{tabular}
  \caption{Results: Afternoon peek (14--18)}
  \label{tab:afternoon_peek}  
\end{table}

\Cref{tab:morning_peek} and \Cref{tab:afternoon_peek} shows the evaluation results for \emph{morning peeks (weekdays, 7--9)} and \emph{afternoon peeks (weekdays, 14--18)} respectively for the time-step $t + 1$.

The peek hour evaluation shows that the \emph{ConvLSTM} model increases its performance over the baseline models when the transport network is put under stress. In the morning peek, the \emph{ConvLSTM} model performs similar as the overall, whereas the baseline models all experience decreased performance of up to several minutes in both the RMSE and MAE, and an increased MAPE of roughly one third.

Likewise the afternoon peek evaluation shows improvements with respect to the baseline models, even though the \emph{ConvLSTM} model also decreases its performance compared to the overall results. The outperformance is neither as significant as in the morning peek, and as the \emph{Google Traffic} model also performs quite well leaving the gap less than a minute.

The get a little more detailed view of how the different models perform on the micro-level (i.e.\ the specific journey) we can inspect a single day of predictions. A random weekday from the test-dataset is plotted in~\Cref{fig:comparison_day}, which shows the accumulated travel time of all 32 links and the predicted travel time at time-step $t + 1$ both for the proposed and baseline models.

In this particular day (a Thursday) the peek hour traffic was worse than normal, which leads both the \emph{historic averaging} model and the \emph{current model} to both underestimate travel time in the peek periods. Remember that the \emph{current model} is based on the \emph{historic averaging} model, so it is not unexpected that they perform very similar. There is also an small peek in travel time in the midday hours, which none of these historic averaging models are able to predict.

On the other hand, both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model get much closer to the ground truth in the peek ours. The \emph{Google Traffic} model seems to predict more accurate then \emph{ConvLSTM} model in the afternoon peek, whereas the opposite occurs in the morning peek. Even more interesting is the detection of the irregular peek in the midday hours, which both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model are able to adjust to, at least to some degree.

\Cref{fig:comparison_day_2}~shows another example (a Friday). Here the difference between the proposed model and the historical averaging and current model baselines are a bit weaker, simply because the the day to a larger degree follows the average pattern for a Friday (especially around the afternoon peek). Even though, the proposed model still performs the best, and it also supports the claim, that the proposed model is strongest when the traffic pattern deviates from the normal pattern, i.e. when the transport network is under stress.

>>>>

\address{
  \vspace{.5em}
  \begin{minipage}{.5\textwidth}
    \center
    Public Transport Movia \\
    Gammel Køge Landevej 3, DK-2500 Valby
  \end{minipage}% This must go next to `\end{minipage}`
  \begin{minipage}{.5\textwidth}
    \center
    Department of Management Engineering \\
    Technical University of Denmark, DK-2800 Kongens Lyngby
  \end{minipage}
  \vspace{-2.5em}
}

\begin{abstract}
Accurate and reliable predictions for travel times in public transport networks is essential for delivering an attractive service.
This paper presents a multi-output, multi-time-step, deep neural network for bus travel time prediction using Convolutional and Long short-term memory (LSTM) layers. The method is empirically evaluated and compared to other popular approaches for link travel time prediction, and currently available services, including the currently deployed model. We find that the proposed model significantly outperforms all the other methods, and is able to detect small irregular peaks in bus travel time very quickly.
\end{abstract}

@@

In this paper we present a multi-output, multi-time-step deep neural network for bus travel time prediction using a combination of \emph{Convolutional} and \emph{Long short-term memory} (LSTM) \cite{Lstm1,Lstm2} layers. The goal of this work is to produce precise short-term predictions (e.g.\ $0-1.5$ hours) for link travel time, specifically for bus traffic in urban areas. The method is empirically evaluated and compared to other popular approaches for link travel time prediction, including the currently model deployed in production by the public transport authority for the Greater Copenhagen Area: \emph{Movia}. Furthermore the method is compared to Google Traffic (part of Google Maps), a popular online service for travel time prediction.

@@

Kalman-filters, which by its capabilities of maintaining state between predictions, has shown interest from several studies for capturing \emph{changes} in the travel time, either as an independent model \cite{Chen2001,Shalaby2004} or in combination with other models \cite{Yo2010,Bai2015}. The Kalman-filter models still have limited options for capturing and forecasting the complex dynamics of travel and dwell time in a metropolitan bus system. E.g., the Kalman-filter's state is only directly accessible for the leading time-step, and thus not capable of finding long-distance patterns spanning over several links and/or over several time-steps. This is substantiated by \cite{Lin2013} and \cite{Kumar2014}, who find artificial neural networks (ANN) to outperform Kalman-filter models. However, the computational challenges of fully connected ANNs are also limiting the depth of the network and thus how complex patterns it can learn to recognize. This has sparked the interest for studying composite or hybrid models. \citet{Bai2015} uses a two-stage approach by combining an offline ANN-model, with an adaptable/online Kalman-filter to yield a dynamic model. The model is not actually learning in the long term, but is able to adapt to temporal variations in the current travel time on a journey, and is still not able to recognize long distance patterns. The proposed model in this work uses \emph{Long short-term memory} cells (LSTM), and differs from existing research in bus link travel time prediction by combining the capability for persisting state-space over multiple time-steps, while allowing the deep neural network to be feasibly trainable.

@@

\section{Convolutional LSTM neural networks}
\label{sec:convlstm}
A Long short-term memory (LSTM) neural network is a special type of Recurrent Neural Network~(RNN) which has proven robust for capturing long-distance dependencies \cite{Lstm1,Lstm2}. The important feature of a LSTM network is its capability to persist cell state, $\matr{c}_t$, from previous observations across sequences of input (e.g.\ time), but also eliminate information which is considered irrelevant. To allow this mechanism the persistence of information is controlled by three gates: \emph{input gate}, \emph{forget gate}, and \emph{output gate}. Each gate yields a state variable at time $t$, respectively $\matr{i}_t$, $\matr{f}_t$, and $\matr{o}_t$, along with the cell output, $\matr{h}_t$, cf.~\cref{eq:lstm}, where $\circ$ denotes the entry-wise product.

\begin{equation}
\begin{aligned}
  \matr{i}_t &= \sigma \left( \matr{W}^i \matr{x}_t + \matr{R}^i \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
  \matr{f}_t &= \sigma \left( \matr{W}^f \matr{x}_t + \matr{R}^f \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c \matr{x}_t + \matr{R}^c \matr{h}_{t-1} + \matr{b}^c \right) \\
  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} \matr{x}_t + \matr{R}^\mathit{o} \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
\end{aligned}
\label{eq:lstm}
\end{equation}
\vspace{.5em}

@@

Convolutional Neural Networks~(CNN) on the other hand have been widely used for capturing spatial relationships, e.g.\ importance of neighboring pixels in an image. As opposed to fully connected layers, where each unit, $i$, in the layer, has a dedicated scalar weight, $w_{ij}$, for all input value, $x_j$, convolutional units are only locally connected and reuse the same weights for producing several outputs. Instead of considering the entire input-vector, only a fixed-size window, or \emph{convolution}, around each input is considered. The weights are therefore referred to as the \emph{filters} or \emph{kernels} of the layer. \Cref{fig:conv} illustrates a single convolutional filter of size $3$ being applied to 1-dimensional data.

@@

\citet{ConvLSTM} introduced the novel combination of Convolutional and LSTM layers into a single structure, the \emph{Convolutional LSTM}, or simply \emph{ConvLSTM}. Specifically the method applies convolutional filters in the \emph{input-to-state} and \emph{state-to-state} transitions of the LSTM cf.~\cref{eq:convlstm}, where $*$ denotes the convolution operator.
\begin{equation}
\begin{aligned}
  \matr{i}_t &= \sigma \left( \matr{W}^i * \matr{x}_t + \matr{R}^i * \matr{h}_{t-1} + \matr{U}^i \circ \matr{c}_{t-1} + \matr{b}^i \right) \\
  \matr{f}_t &= \sigma \left( \matr{W}^f * \matr{x}_t + \matr{R}^f * \matr{h}_{t-1} + \matr{U}^f \circ \matr{c}_{t-1} + \matr{b}^f \right) \\
  \matr{c}_t &= \matr{f}_t \circ \matr{c}_{t-1} + \matr{i}_t \circ \mathrm{tanh} \left( \matr{W}^c * \matr{x}_t + \matr{R}^c * \matr{h}_{t-1} + \matr{b}^c \right) \\
  \matr{o}_t &= \sigma \left( \matr{W}^\mathit{o} * \matr{x}_t + \matr{R}^\mathit{o} * \matr{h}_{t-1} + \matr{U}^\mathit{o} \circ \matr{c}_{t} + \matr{b}^o \right) \\
  \matr{h}_t &= \matr{o}_t \circ \mathrm{tanh} \left( \matr{c}_t \right)
\end{aligned}
\label{eq:convlstm}
\end{equation}
\vspace{.5em}

@@

Thus convolutional filters are applied to each input, at each time-step, to the respective LSTM-cell, and also between LSTM-cells in the state-transition. Since the time-steps are one-dimensional~(i.e.\ link travel times across links), the filters are also one-dimensional. In each of the two blocks, the \emph{ConvLSTMs} are arranged with filter sizes of respectively $10\times1$ and $5\times1$ for each of the layers in the block. This size is used both for the \emph{input-to-state} and \emph{state-to-state} convolutional filters. Finally each \emph{ConvLSTM} layer has 64~outputs, yielding a total need of 512 convolutional filters.

@@

For the \emph{ConvLSTM} model to be able to capture the desired spatial--temporal patterns, the input data must be arranged in a suitable manner: I.e.\ in $N$ samples, each with a window of the $w$ lagging time-steps $t-w+1, \ldots, t$, and each time-step with $u$ link travel times $1, \ldots, u$ cf.~\Cref{fig:data_shape}.

@@

\subsection{Detrending}
Urban bus travel times vary throughout the time of the day, and the day of the week due to \emph{recurring congestion}. In order to reduce the need for the network to learn this recurring variation, link travel time for link $\mathit{ln} \in \{ 1,\ldots,u\}$, at time-step $t$, $x_{\mathit{ln},t}$, is normalized to focus on deviations from the normal and expected pattern. Travel times are centered with the mean for each link, at the time of day, and day of week, $\mathit{\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}}$, and scaled with the standard deviation for each link, $\sigma_\mathit{ln}$, cf.~\cref{eq:normalization}.

@@

Given the time resolution, we set the fixed window size, $w = 32$, equivalent of 8 hours, to allow pattern in the morning peak affect patterns in the afternoon peak. We set $k = 3$ to allow predictions of up to 45 minutes into the future.

@@

\section{Results and discussion}
\label{sec:results}
The performance of our proposed model, based on \emph{ConvLSTM}, for link travel time prediction is compared against several other baseline models and services using the presented evaluation:
\begin{enumerate}
   \item a naïve historical average model, i.e.\ equivalent of just predicting the normalized value, $\bar{x}_{\mathit{ln},\mathit{dow},\mathit{tod}}$ 
   \item the traffic prediction model currently deployed by Movia.
   \item a pure LSTM-model for link travel time prediction, i.e.\ without applying convolutional filters in state transitions.
   \item travel time predictions from Google Traffic (part of Google Maps).
\end{enumerate}

@@

\subsection{Current model}
Measurements for the currently deployed bus prediction model were collected at a 5-minute frequency using a non-publicly accessible endpoint at the transport authority. The model is based on an historical average model, but also has a rule based mechanism on top, which can override or adjust the historical link travel times. For instance it will assume that a delayed vehicle can recover (partially or completely) from its delay by traversing links faster. Of course such an assumption can be problematic in an urban area with lots of external traffic effects.

@@

\subsection{Google Traffic}
Measurements from the Google Traffic model were collected using the Google Maps Distance Matrix API~\citep{GMaps_DistanceMatrixAPI}. Google uses crowd-sourced road congestion data collected from smart-phones with the \emph{Google Maps App} installed \citep{GMaps_Crowdsourcing}. While the exact model powering the service is not publicly described in detail, the documentation states that ``the returned \emph{duration in traffic} should be the best estimate of travel time given what is known about both historical traffic conditions and live traffic''. And further, ``that live traffic becomes more important the closer the departure time is to now'' \citep{GMaps_DistanceMatrixAPI}.

@@

Another important aspect is that the Google Traffic model is primary designed for estimating car travel times, and can thus be biased away from the bus travel times used in this experiment. Since we only consider link travel time, and collects data for each link individually, the bus dwell time will not be an issue, as it is not included in either measurement.

@@

Although the advantage might seem small it is emphasized that evaluation measurements are averaging their response, and thus the increased accuracy can be much higher on individual journeys, especially if they experience very irregular travel times. To investigate this, we focus our analysis on periods when the transport system is most vulnerable, and even small changes in regularity can propagate, since recovery is not an option, i.e.\ morning and afternoon peaks.

\begin{table}[!ht]
  \center
  \begin{tabular}{l|rrr}
  Model &  RMSE &  MAE &  MAPE \\
  \hline
  \hline
  Historical Average &        6.40 &       5.57 &     10.62 \% \\ \hline
  Current Model      &        6.69 &       5.88 &     11.22 \% \\ \hline
  Pure LSTM          &        0.00 &       0.00 &      0.00 \% \\ \hline
  Google Traffic     &        5.25 &       4.62 &      9.17 \% \\ \hline
  ConvLSTM           &        2.64 &       2.09 &      4.04 \% \\ \hline
  \end{tabular}
  \caption{Results: Morning peak (7--9)}
  \label{tab:morning_peak}

@@

\begin{table}[!ht]
  \center
  \begin{tabular}{l|rrr}
  Model &  RMSE &  MAE &  MAPE \\
  \hline
  \hline
  Historical Average &    5.90 &   4.65 &  8.28 \% \\ \hline
  Current Model      &    6.28 &   5.20 &  9.37 \% \\ \hline
  Pure LSTM          &    0.00 &   0.00 &  0.00 \% \\ \hline
  Google Traffic     &    4.16 &   3.34 &  6.21 \% \\ \hline
  ConvLSTM           &    3.79 &   3.02 &  5.61 \% \\ \hline
  \end{tabular}
  \caption{Results: Afternoon peak (14--18)}
  \label{tab:afternoon_peak}  
\end{table}

\Cref{tab:morning_peak} and \Cref{tab:afternoon_peak} shows the evaluation results for \emph{morning peaks (weekdays, 7--9)} and \emph{afternoon peaks (weekdays, 14--18)} respectively for the time-step $t + 1$.

The peak hour evaluation shows that the \emph{ConvLSTM} model increases its performance over the baseline models when the transport network is put under stress. In the morning peak, the \emph{ConvLSTM} model performs similar as the overall, whereas the baseline models all experience decreased performance of up to several minutes in both the RMSE and MAE, and an increased MAPE of roughly one third.

Likewise the afternoon peak evaluation shows improvements with respect to the baseline models, even though the \emph{ConvLSTM} model also decreases its performance compared to the overall results. The outperformance is neither as significant as in the morning peak, and as the \emph{Google Traffic} model also performs quite well leaving the gap less than a minute.

To get a more detailed view of how the different models perform on the micro-level (i.e.\ the specific journey) we can inspect a single day of predictions. A random weekday from the test-dataset is plotted in~\Cref{fig:comparison_day}, which shows the accumulated travel time of all 32 links and the predicted travel time at time-step $t + 1$ both for the proposed and baseline models.

In this particular day (a Thursday) the peak hour traffic was worse than normal, which leads both the \emph{historic averaging} model and the \emph{current model} to both underestimate travel time in the peak periods. Remember that the \emph{current model} is based on the \emph{historic averaging} model, so it is not unexpected that they perform very similar. There is also small peak in travel time in the midday hours, which none of these historic averaging models are able to predict.

On the other hand, both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model get much closer to the ground truth in the peak hours. The \emph{Google Traffic} model seems to predict more accurate then \emph{ConvLSTM} model in the afternoon peak, whereas the opposite occurs in the morning peak. Even more interesting is the detection of the irregular peak in the midday hours, which both the \emph{Google Traffic} model and the proposed \emph{ConvLSTM} model are able to adjust to, at least to some degree.

\Cref{fig:comparison_day_2}~shows another example (a Friday). Here the difference between the proposed model and the historical averaging and current model baselines are a bit weaker, simply because the the day to a larger degree follows the average pattern for a Friday (especially around the afternoon peak). Even though, the proposed model still performs the best, and this also supports the claim, that the proposed model is strongest when the traffic pattern deviates from the normal pattern, i.e.\ when the transport network is under stress.
